---
title: 'Lecture 7: Modeling'
output:
  html_notebook: default
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
---

You have probably seen a line written as y = mx + b. 
Also, you might have seen Beer's Law before in a biology or chemistry class. 
```
Beer's law: absorbance = constant * concentration 
```
It basically just says that absorbance varies linearly with concentration. Y = absorbance, m = constant, and x = concentration in this case. 

To learn about modeling we'll be using the R built in dataset called Formaldehyde. Learn more about this dataset by running this block of code:
```{r}
?Formaldehyde
View(Formaldehyde)
```

Formaldehye has 2 columns

1. carb = numeric	Carbohydrate (ml), basically concentration of sample

2. optden =	numeric	Optical Density, absorbance of sample


From Beer's law, we expect a graph of optden vs carb to form a line.
Let's test out that hypothesis using the plot() function, which plots y vs x. 

```{r}
plot(x = Formaldehyde$carb, y = Formaldehyde$optden)
```

This looks pretty linear to me!

\newline

We can add a title, x-axis label, and y-axis label with additional parameters

* main = title
* xlab = x axis label
* ylab = y axis label

```{r}
plot(x = Formaldehyde$carb, y = Formaldehyde$optden, main = "Beer's Law", xlab = "concentration", ylab = "absorbance")
```

Just visually seeing a linear shape unfortunately isn't enough evidence to say there is a linear relationship. We can actually fit a straight line through our data. The function that creates a linear model is called **lm()**. We pass in our y and x variables with a tilda in the middle. This following code stores that linear model into a variable valled _linear_model_.

```{r}
linear_model <- lm(Formaldehyde$optden ~ Formaldehyde$carb)
```

To see what the actual linear model is, we can print it using **print()**

```{r}
print(linear_model)
```

This tells us the y-intercept (0.005086) and slope (0.876286) for our linear model. 

\newline


What would give us even more information is calling **summary()** (which we have learned before for vectors) on our model. 
 
```{r}
summary(linear_model)
```
This tells both the y-intercept and slope that we saw before, but it also contains important statistics such as standard error and p-values (we'll discuss p-values in our next lecture).

\newline

Now that we have our model, let's add it to our plot! The function is abline(), and we pass the variable storing our model in as a parameter
```{r}
plot(x = Formaldehyde$carb, y = Formaldehyde$optden, main = "Beer's Law", xlab = "concentration", ylab = "absorbance")
abline(linear_model)
```

Here are the solutions to the in-class exercises:
```{r}
conc_liters <- Formaldehyde$carb * 0.001

plot(x = conc_liters, y = Formaldehyde$optden, 
     main = "Beer's Law", 
     xlab = "concentration (L)", 
     ylab = "absorbance")

linear_model_liters <- lm(Formaldehyde$optden ~ conc_liters)
summary(linear_model_liters)

abline(linear_model_liters)

```




Let's see another example that combines information from our data cleaning lecture. 
```{r}
#Import the dataset and omit the NA's 
complete_heart <- na.omit(read.csv("C:/Users/miche/Desktop/STIC/Read File Examples/heart_missing.csv", na.strings = "NA"))
nrow(complete_heart)

#Plot BMI vs chol
plot(x = complete_heart$chol, y = complete_heart$BMI)
linear_model <- lm(complete_heart$BMI ~ complete_heart$chol)
abline(linear_model)

#Does smoking modulate the relationship between cholesterol and BMI?
non_smokers <- subset(complete_heart, complete_heart$smoking == 0)
smokers <- subset(complete_heart, complete_heart$smoking == 1)

plot(x = smokers$chol, y = smokers$BMI)
plot(x = non_smokers$chol, y = non_smokers$BMI)

#We can compare the slopes and intercepts of these two lines
#Model stratifies by smoking status
model_smoker <- lm(smokers$BMI ~ smokers$chol)
summary(model_smoker)
model_non_smoker <- lm(non_smokers$BMI ~ non_smokers$chol)
summary(model_non_smoker)

#Help visualize both lines 
plot(x = complete_heart$chol, y = complete_heart$BMI)
abline(model_smoker)
abline(model_non_smoker)
```


We learned the lm() function to create linear models. There are TONS of models out there, such as logistic regression, exponential regression, polynomial regression, etc. 

\newline

We won't be doing too much with these other types of regression because I understand everyone is coming to this with a different level of statistics background. However, know that these exist, and modeling with another regression follows this general sort of principle. 
\newline
For example, logistic regression would look like this. The function **lm()** is replaced by **glm()**, which stands for generalized linear model. We also add an extra parameter at the end to specify distribution as binomial. 

```
logistic_model <- glm(y_variable ~ x_variable, family = "binomial")
```

Again, don't worry about this if you don't know what logistic regression is, but if you were interested in other types of models, **glm()** pretty much has you covered. 

```{r}
?glm
```

